---
title: "Getting Started with Rpsge"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with Rpsge}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(Rpsge)
```

## Introduction

Probabilistic Structured Grammatical Evolution (PSGE) is a grammar-guided genetic programming approach that combines the benefits of structured representation with probabilistic grammar learning. This vignette demonstrates how to use the Rpsge package to solve problems using PSGE.

## Installation

You can install Rpsge from GitHub:
  
```r
# install.packages("devtools")
devtools::install_github("sjvrensburg/Rpsge")
```

## Basic Concepts

PSGE works by evolving solutions within the constraints of a context-free grammar. The key components are:
  
1. **Grammar**: Defines the structure and syntax of possible solutions
2. **Genotype**: A structured representation of real values between 0 and 1
3. **Phenotype**: The actual solution derived from mapping the genotype using the grammar
4. **Fitness Function**: Evaluates how well a solution performs

## Creating a Grammar

Grammars are defined in BNF (Backus-Naur Form) notation. Here's an example grammar for symbolic regression:

```bnf
<expr> ::= <expr> <op> <expr> | <var>
<op> ::= + | - | * | /
<var> ::= x | 1.0
```

Save this in a file (e.g., `symbolic.bnf`) and load it:

```{r eval=FALSE}
library(Rpsge)

grammar <- read_pcfg_grammar("symbolic.bnf")
print_grammar(grammar)
```

## Defining a Fitness Function

The fitness function evaluates how well a solution performs. Here's an example for symbolic regression:
  
```{r eval=FALSE}
fitness_fn <- function(phenotype) {
  # Target function: x^2 + x + 1
  f <- function(x) {
    tryCatch({
      eval(parse(text = phenotype))
    }, error = function(e) {
      return(Inf)  # Return Inf for invalid expressions
    })
  }
  
  # Generate test points
  x <- seq(-1, 1, by = 0.1)
  y_true <- x^2 + x + 1
  
  # Evaluate solution
  y_pred <- sapply(x, f)
  
  # Handle invalid results
  if (any(is.infinite(y_pred)) || any(is.nan(y_pred))) {
    return(Inf)
  }
  
  # Return RMSE
  sqrt(mean((y_pred - y_true)^2))
}
```

## Running PSGE

PSGE includes progress reporting through the progressr package:
  
```{r eval=FALSE}
library(progressr)

# Configure progress reporting
handlers(global = TRUE)
handlers("progress")

# Run PSGE with progress reporting
with_progress({
  result <- psge(
    grammar_file = "symbolic.bnf",
    fitness_fn = fitness_fn,
    pop_size = 100,
    generations = 50,
    crossover_rate = 0.9,
    mutation_rate = 0.1,
    tournament_size = 3,
    max_depth = 10,
    learning_factor = 0.01
  )
})
```

## Analyzing Results

The `result` object contains:
  
```{r eval=FALSE}
# Best solution found
print(result$best_solution$phenotype)
print(result$best_solution$fitness)

# Evolution history
head(result$history)

# Plot evolution progress
plot_evolution(result$history)

# Final grammar with learned probabilities
print_grammar(result$final_grammar)
```

## Parallel Evaluation

For computationally intensive fitness functions, you can enable parallel evaluation:
  
```{r eval=FALSE}
with_progress({
  result <- psge(
    grammar_file = "symbolic.bnf",
    fitness_fn = fitness_fn,
    parallel = TRUE,
    n_workers = 4  # Number of parallel workers
  )
})
```

## Advanced Features

### Grammar Analysis

```{r eval=FALSE}
# Calculate minimum depths required for each non-terminal
depths <- calculate_min_depths(grammar)

# Analyze grammar usage in population
usage_stats <- analyze_grammar_usage(grammar, result$final_population)
print(usage_stats)

# Visualize grammar structure
library(DiagrammeR)
visualize_grammar(grammar, max_depth = 3)
```

### Custom Post-Generation Functions

You can provide custom functions to run after each generation or evaluation:
  
```{r eval=FALSE}
# Example: Track unique phenotypes
diversity_history <- data.frame()

post_eval_fn <- function(population, context) {
  # Calculate diversity
  unique_count <- length(unique(sapply(population, function(ind) ind$phenotype)))
  
  # Store in our history
  diversity_history <<- rbind(
    diversity_history,
    data.frame(
      generation = context$generation,
      unique_phenotypes = unique_count
    )
  )
  
  return(population)
}

# Use in PSGE
result <- psge(
  grammar_file = "symbolic.bnf",
  fitness_fn = fitness_fn,
  post_evaluation_fn = post_eval_fn
)
```

### Sampling from Grammar

You can generate random samples from your grammar:
  
```{r eval=FALSE}
# Generate 10 random phenotypes
samples <- sample_phenotypes(grammar, n = 10)
print(samples)
```

## Tips for Effective Use

1. **Grammar Design**
  - Keep grammars as simple as possible while capturing the problem structure
- Include both recursive and non-recursive rules for each non-terminal
- Consider the minimum depth required for valid solutions

2. **Parameter Tuning**
  - Start with default parameters
- Increase population size for complex problems
- Adjust tournament size to control selection pressure
- Tune learning factor to balance exploitation and exploration

3. **Fitness Function Design**
  - Handle invalid solutions gracefully
- Consider computational efficiency
- Normalize fitness values if possible
- Use parallel evaluation for expensive computations

## Common Issues and Solutions

1. **Invalid Solutions**
```r
# Handle invalid phenotypes in fitness function
fitness_fn <- function(phenotype) {
  result <- tryCatch({
    # Your evaluation code
  }, error = function(e) {
    return(Inf)  # Or other penalty value
  })
  return(result)
}
```

2. **Premature Convergence**
```r
# Increase population diversity
result <- psge(
  grammar_file = "symbolic.bnf",
  fitness_fn = fitness_fn,
  tournament_size = 2,  # Reduce selection pressure
  mutation_rate = 0.2,  # Increase exploration
  learning_factor = 0.005  # Slower probability adaptation
)
```

3. **Slow Evolution**
```r
# Enable parallel evaluation
result <- psge(
  grammar_file = "symbolic.bnf",
  fitness_fn = fitness_fn,
  parallel = TRUE,
  n_workers = 4,
  pop_size = 200  # Larger population
)
```

## Conclusion

PSGE combines the structure of grammar-guided evolution with probabilistic learning to create an effective approach to genetic programming. The Rpsge package provides a comprehensive implementation with features for analysis, visualization, and performance optimization.

## References

```bibtex
@inproceedings{megane2022psge,
  title={Probabilistic Structured Grammatical Evolution},
  author={M{\'e}gane, Jessica and Louren{\c{c}}o, Nuno and Machado, Penousal},
  booktitle={2022 IEEE Congress on Evolutionary Computation (CEC)},
  pages={1--9},
  year={2022},
  doi={10.1109/CEC55065.2022.9870397}
}
```
